Data Migration under a Changing Schema

### 2023-07-01 2022Migration/articleMigration.tex
Work on the process description (new chp 4)

\subsection{Process of Data Migration}

@@

   Let us first sketch the steps to migrate system $\infsys$ to $\infsys'$.

@@

   \item Let the existing information system $\infsys$ have data set $\dataset$ and schema $\schema$.
         A developer defines a new information system, $\infsys'$, which contains a data set of its own, $\dataset'$, a schema $\schema'$, and some functionality\footnote{In this paper, we only consider the functionality captured by rules.} as illustrated in figure~\ref{fig:post-migration}.
         After the migration, the resulting system will be $\infsys'$, enhanced with the triples from $\dataset$ that satisfy $\schema'$,
         and with additional changes made by the migration engineer (see step~\ref{item3}) and by users who have been using the system during the migration.
         $\dataset'$ may contain some new data,
         for instance, to initialize new features at the moment of transition (MoT).
         If schemas $\schema$ equals $\schema'$, the data migration is unnecessary, so in the sequel we assume they are different.
   \item The migration engineer uses a software generator to generate a migration script, $\migrsys$,
         This is a system that includes both $\infsys$ and $\infsys'$.
         $\migrsys$ is meant to migrate the data and define the semantics during the migration phase.
         It contains enforcement rules that migrate existing data to the desired data set.
         It also contains data structures and rules that are needed for the duration of the migration phase.
%         The generated script must be free of type errors, to ensure that $\sat{u}{\dataset}$ exists for every rule $u$ and data set $\dataset$.
%         The proof is available \href{https://www.isa-afp.org/}{here}

@@

         In many cases, the schema $\schema_{\migrsys}$ will be enough to perform the migration.

@@

         For that purpose, the generator generates the migration script as an Ampersand script,
         so the migration engineer can accommodate such requirements.

@@

         She can either implement enforcement rules or use the business to resolve the issues.

@@

         The migration engineer alters $\schema$ nor $\schema'$ while adapting $\migrsys'$ to the specific needs of the current migration.
   \item Then, the migration engineer deploys system $\migrsys'$.
         This event starts the copying of data from $\dataset$ to $\dataset'$, as specified in $\migrsys'$ (figure~\ref{fig:migration phase}).

@@

         Users are still changing $\dataset$ (both inserts and deletes),
         but these changes are being transferred to $\dataset'$ as well.

@@

   \item At this moment, the ingress switches the incoming traffic from $\infsys$ to $\migrsys'$,

@@

@@

         After that, dataset $\dataset$ will no longer change.
   \item From this moment, users experience the functionality of the migration system $\migrsys'$.
         This system must ensure that all invariants of $\schema'$ are true before the migration is complete.
         It no longer ensures the rules of $\schema$, which govern the existing system.
         Since $\dataset$ is no longer changing, the migration engineer can now focus on the remaining violations of $\schema'$.
         Users of the migration system will deal with process rules in the migration system that are still violated.

@@

         This state marks the moment of cleanup (MoC)
   \item After the MoC, the ingress switches the incoming traffic from $\migrsys'$ to $\infsys'$.
         The migration system is no longer needed and can be taken out of production, together with the existing system $\infsys$.
         The desired system remains in production and the migration is finished.

@@

   Upon initialization, the migration system evaluates all enforce rules in an attempt to satisfy all invariants.
   In this example, it leaves two violations behind that it cannot resolve by enforce rules.
   The pair {\tt ("Peter", "Management")} is the first one; it violates rule {\tt new.ExamRule2}.
   This is because Peter has registered for an examination of the course Management,
   which contains the module Finance for which Peter is not enrolled.
   The other is {\tt "IT-Governance"}, which violates the univalence of relation {\tt new.isPartOf}.
   Since rule {\tt new.ExamRule2} and the univalence of {\tt new.isPartOf} are both defined as invariants,
   the desired system cannot be taken into production until these violations are resolved.
   Until that moment, the migration system remains operational to allow a user to resolve this violation. 
   The moment of cleanup is marked by the event that all invariants of the Migration system and the desired system are satisfied.
   At that moment, the migration system is no longer needed and can be removed.

@@

When the default migration system $\migrsys$ is deployed, the system will always be in a state such that continually running any violated ENFORCE rules (in any order) leads to the MoT.
   This means that $A\rightsquigarrow B$ is such that the dataset of $B$ is that of $A$ with an update applied according to one or more violated enforce rules. % TODO: define 'violated'
   If there is no such $B$ because no enforce rules are violated, we are at the moment of transition.
   We say that the system $A$ is in a state such that the MoT can be reached if there are no infinite $\rightsquigarrow$-paths from $A$.
   The constraint on $A\dashrightarrow B$ is that only the old system can change at this point.
   Note that this does not prove that the moment of transition to never be reached, as more enforce rules may get violated at every $\dashrightarrow$ step.
   We expect this not to be a problem in practice since $\dashrightarrow$ transitions are triggered by users while $\rightsquigarrow$ is triggered automatically through ENFORCE rules: the ENFORCE rules should outperform the $\dashrightarrow$ actions by a big margin.

@@

\section{Software architecture}
   The theory presented above roughly corresponds to the back-end of a migration system.
   This section describes its software architecture.
   For this research, we have used the Ampersand compiler to generate the system(s).
   Figure~\ref{fig:initial} shows the starting point of the migration.
\begin{figure}
   \centering
   \includegraphics[scale=0.1]{figures/Before migration.png}
   \caption{Initial state of the migration}
   \label{fig:initial}
\end{figure}
   For the purpose of data migration, the existing system must make its schema available through its API.
   For this purpose, the back-end offers a service ``retrieve schema'', which provides the schema as Ampersand source-code.
   As a result, the migration engineer can rely on the schema of any generated application to be accurate.
   For this reason, we can safely assume that the schema this service provides has not been changed since it was deployed.
   Its presence allows the generator to build various front-ends and migration scripts during the lifetime of the existing system%

   After the migration engineer has produced and tested a migration script, the Ampersand compiler produces a migration system.
   It includes the schemas of both the existing system and the desired system, besides having a schema of its own.
   This is implemented by Ampersand's INCLUDE mechanism, which boils down to taking the disjoint union of the schemas.
\begin{figure}
   \centering
   \includegraphics[scale=0.1]{figures/During migration.png}
   \caption{During the migration}
   \label{fig:during}
\end{figure}
   Deployment of the migration system (figure~\ref{fig:during}) leaves the existing system intact.
   It enhances the system with the data set needed by the migration system, which includes the data set needed by the desired system.
   So, users will notice no difference in the existing system when the migration system is deployed.

>>>>

\section{Process of Data Migration}

@@

\begin{figure}
   \centering
   \includegraphics[scale=0.1]{figures/Before migration.png}
   \caption{Initial state of the migration}
   \label{fig:initial}
\end{figure}
   Figure~\ref{fig:initial} shows the initial state of the migration,
   in which the existing information system $\infsys$ has data set $\dataset$ and schema $\schema$.
   At the start of this migration, $\dataset$ satisfies all rules of $\schema$ (by requirement~\ref{eqn:satisfaction}).

@@

\item The migration starts with a developer%
\footnote{The developer role may stand for a team of developers, a single developer, a migration engineer, or whoever defines the desired system.}
      who specifies a new information system, $\infsys'$.
      Let us call this moment the start of the migration (SoM).
      $\infsys'$ contains a data set of its own, $\dataset'$, a schema $\schema'$, and some functionality%
\footnote{In this paper, we only consider functionality that is captured by rules.}.
      Before replacing $\infsys$ by $\infsys'$, the data from $\dataset$ must be migrated to $\dataset'$
      and all invariants of $\schema'$ must be satisfied.
      We assume that schemas $\schema$ and $\schema'$ are different, or else the migration would be unnecessary.
\item From schemas $\schema$ and $\schema'$, a migration engineer%
\footnote{The migration engineer may also stand for a team, which may overlap with the developer role.}
      generates a migration script, using the Ampersand compiler.
      This script specifies the migration system, $\migrsys$, that includes $\infsys$, $\infsys'$ and a number of migration rules.
      $\migrsys$ is meant to migrate the data and ensure that $\dataset'$ satisfies all invariants of $\schema'$.

@@

      In specific cases, the schema $\schema_{\migrsys}$ will be enough to perform the migration automatically.
      As soon as the data is transferred and $\migrsys$ has satisfied the invariants of $\schema'$,
      $\infsys'$ can take care of production on its own without the help of $\migrsys$.

@@

      For that purpose, she may want to alter the migration script to accommodate such requirements.

@@

      She can implement such adaptations as enforcement rules, which are executed automatically.
      But she can also ask business users to resolve issues by specifying process rules.
      Human involvement is often needed to implement a desired migration strategy.
      In that case the migration system must ensure that there are finitely many steps to complete the migration.
      The migration engineer must ensure that $\migrsys'$ can satisfy all invariants in $\schema'$, either by enforcement rules or by process rules.

@@

      When this work is done, the migration engineer has a migration system $\migrsys'$ with schema $\schema'$ and an initial dataset $\dataset'$
      that is ready to deploy.
\item\label{item4}
      Then, the migration engineer deploys system $\migrsys'$.
\begin{figure}
   \centering
   \includegraphics[scale=0.1]{figures/4MoD.png}
   \caption{During the migration}
   \label{fig:4MoD}
\end{figure}
      Let us call this moment the moment of deployment (MoD).
      This event starts the copying of data from $\dataset$ to $\dataset'$, as specified in $\migrsys'$.

@@

      During this phase, users can still change $\dataset$ (both inserts and deletes) while $\infsys$ is keeping its invariants true.
      Yet, the migration system also transfers these ``new'' changes to $\dataset'$ by means of its enforcement rules.

@@

\item After completing step~\ref{item4}, the ingress switches the incoming traffic from $\infsys$ to $\migrsys'$,

@@

\begin{figure}
   \centering
   \includegraphics[scale=0.1]{figures/5MoT.png}
   \caption{Situation after MoT}
   \label{fig:during}
\end{figure}
      Simultaneously, all clients must switch to the functionality of $\migrsys'$,
      so users will experience the new functionality.

@@

      After that, the existing dataset $\dataset$ will no longer change.
\item Starting at MoT, $\migrsys'$ is working towards satisfying the migration rules.
      It executes its enforce rules to migrate data.
      It also executes enforce rules and process rules to ensure that all invariants of $\infsys'$ are true.
      Since $\migrsys'$ also contains $\schema$,
      it will also keep the invariants of $\infsys$ as the final transactions trickle into the new system.
      However, the traffic into $\infsys$ has stopped, so the changes in $\dataset$ will quickly dry up.

@@

\item Let us designate the moment of completion as MoC.
      After the MoC, the ingress can switch the incoming traffic from $\migrsys'$ to $\infsys'$ because all invariants of $\infsys'$ are true.
      $\infsys$ and $\migrsys'$ are no longer needed and can be taken out of production.
      $\infsys'$ remains in production and the migration is finished.

@@

   Upon initialization, the migration system evaluates all enforcement rules to satisfy all invariants.
   In this example, it leaves two violations behind that it cannot resolve by enforcement rules.
   The pair {\tt ("Peter", "Management")} is the first one; it violates rule {\tt new.ExamRule2}.
   This is because Peter has registered for an examination of the course Management,
   which contains the module Finance for which Peter is not enrolled.
   The other is {\tt "IT-Governance"}, which violates the univalence of relation {\tt new.isPartOf}.
   Since rule {\tt new.ExamRule2} and the univalence of {\tt new.isPartOf} are both defined as invariants,
   the desired system cannot be taken into production until these violations are resolved.
   Until that moment, the migration system remains operational to allow a user to resolve this violation. 
   The moment of cleanup is marked by the event that all invariants of the Migration system and the desired system are satisfied.
   At that moment, the migration system is no longer needed and can be removed.

@@

When the default migration system $\migrsys$ is deployed, the system will always be in a state such that continually running any violated ENFORCE rules (in any order) leads to the MoT.
   This means that $A\rightsquigarrow B$ is such that the dataset of $B$ is that of $A$ with an update applied according to one or more violated enforcement rules. % TODO: define 'violated'
   If there is no such $B$ because no enforcement rules are violated, we are at the moment of transition.
   We say that the system $A$ is in a state such that the MoT can be reached if there are no infinite $\rightsquigarrow$-paths from $A$.
   The constraint on $A\dashrightarrow B$ is that only the old system can change at this point.
   Note that this does not prove that the moment of transition to never be reached, as more enforcement rules may get violated at every $\dashrightarrow$ step.
   We expect this not to be a problem in practice since $\dashrightarrow$ transitions are triggered by users while $\rightsquigarrow$ is triggered automatically through ENFORCE rules: the ENFORCE rules should outperform the $\dashrightarrow$ actions by a big margin.

@@

\section{Software architecture}
   The theory presented above roughly corresponds to the back-end of a migration system.
   This section describes its software architecture.
   For this research, we have used the Ampersand compiler to generate the system(s).
   For the purpose of data migration, the existing system must make its schema available through its API.
   For this purpose, the back-end offers a service ``retrieve schema'', which provides the schema as Ampersand source-code.
   As a result, the migration engineer can rely on the schema of any generated application to be accurate.
   For this reason, we can safely assume that the schema this service provides has not been changed since it was deployed.
   Its presence allows the generator to build various front-ends and migration scripts during the lifetime of the existing system%

   After the migration engineer has produced and tested a migration script, the Ampersand compiler produces a migration system.
   It includes the schemas of both the existing system and the desired system, besides having a schema of its own.
   This is implemented by Ampersand's INCLUDE mechanism, which boils down to taking the disjoint union of the schemas.
   Deployment of the migration system leaves the existing system intact.
   It enhances the system with the data set needed by the migration system, which includes the data set needed by the desired system.
   So, users will notice no difference in the existing system when the migration system is deployed.
